k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-02-21 paper.Rmd
minor stuff abstract

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise within-set dissimilarity ). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an integer linear programming algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. Two simulation studies provide an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so. A practical example shows how the extended BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.

>>>>

    When assigning stimuli to sets in experimental psychology, researchers strive for high similarity between stimulus sets and high pairwise dissimilarity of stimuli within sets. Because stimulus assignment is usually inadequate when done manually, automated algorithmic approaches are preferred. Brusco et al. (2020) proposed a bicriterion anticlustering algorithm (BILS) that simultaneously maximizes the diversity (corresponding to between-set similarity) and the dispersion (corresponding to pairwise within-set dissimilarity). We provide a first independent assessment of BILS and extend it in three ways. First, by prefixing an exact algorithm, our extension ensures that a globally optimal value of dispersion is always returned; without it, it is unknown if the BILS's result is optimal. Second, the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, for example to prevent specific pairs of stimuli from appearing in the same condition. Third, our extension allows that other anticlustering objectives of between-group similarity -- such as the k-means or k-plus objectives -- can be optimized instead of only the diversity. A simulation study provides an extensive evaluation of the original BILS and our extension. Using the optimal approach as benchmark, we show that the original BILS identifies optimal dispersion values for much larger data sets than previously analyzed, but we also highlight conditions where it fails to do so. A practical example shows how the extended BILS can be used for stimulus assignment in experimental psychology. The original BILS as well as our extension are both freely available via the open source R package anticlust.
